---
title: "Class 12"
author: "Isabel Hui - A16887852"
format: pdf
---

## Section 1: Proportion of G/G in population

Downloaded a CSV file from Ensemble <https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39895045-39895146;v=rs8067378;vdb=variation;vf=959672880;sample=HG00109>

Now we can read this CSV file.

```{r}
mxl <- read.csv("samplegenotypes.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

## Section 4: Population Scale Analysis [HOMEWORK]

One sample is not enough to know what is happening in a population.

How many samples do we have?

```{r}
expr <- read.table("rs8067378.txt")
head(expr)
```

```{r}
nrow(expr)
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.


```{r}
table(expr$geno)
```

```{r}
library(dplyr)
df <- expr
str(df)
```

```{r}
summary_df <- df %>%
  group_by(geno) %>%
  summarise(
    sample_size = n(),  
    median_expression = median(exp, na.rm = TRUE)
  )

summary_df
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

The box plot does show differing levels of expression between the A/A and G/G genotype. The SNP does effect the expression of ORMDL3, as a G/G genotype (blue box) can be seen to significantly decrease expression as the mean and top and bottom quartile all lie below the values of the A/A genotype expression.



