---
title: "Class 10: Halloween Mini-Project"
author: "Isabel Hui - A16887852"
format: gfm
---

Today is Halloween, an old Irish holiday. Let's celebrate by eating candy.

We will explore some data all about Halloween candy from the 538 website.

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy <- read.csv(candy_file, row.names = 1)

head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

```{r}
rownames(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

```{r}
sum(candy$chocolate)
```


> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Milky Way Midnight", ]$winpercent
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```


```{r}
library(dplyr)

candy |>
  filter(rownames(candy)=="Haribo Happy Cola") |>
  select(winpercent)
```

> Q. Find fruity candy with a winpercent above 50%.

```{r}
candy |>
  filter(winpercent > 50) |>
  filter(fruity==1)
```

To get a quick insight into a new dataset, some people like using the skimer package and its `skim()` function.

```{r}
library(skimr)
skimr::skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Looks like the `winpercent` variable or column is measured on a different scale than everything else! I will need to scale my data before doing any analysis like PCA etc.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

They represent True/False.

> Q8. Plot a histogram of winpercent values.

We can do this a few ways, e.g. the "base" R `hist()` function or with `ggplot()`.
```{r}
hist(candy$winpercent, breaks=10)
```
```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth = 8) +
  theme_minimal()
```


> Q9. Is the distribution of winpercent values symmetrical?

No.

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
fruit.candy <- candy |>
  filter(fruity == 1)

summary(fruit.candy$winpercent)
```

```{r}
#summary(candy[as.logical(candy$chocolate),]$winpercent)

choc.candy <- candy |>
  filter(chocolate == 1)

summary(choc.candy$winpercent)
```

Chocolate candy has a higher median and mean compared to fruit candy.

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.candy$winpercent, fruit.candy$winpercent)
```

> Q13. What are the five least liked candy types in this set?



> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
play <- c("d", "a", "c")
sort(play)
order(play)
```

```{r}
sort(c(5,2,10), decreasing = T)
```


```{r}
head(candy[order(candy$winpercent),], 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

Let's do a barplot of winpercent values.

```{r}
ggplot(candy) +
  aes(x = winpercent, 
      y = reorder(rownames(candy),winpercent),
      fill = chocolate) +
  geom_col()
```

I want a more custom color scheme where I can see both chocolate and bar and fruity etc. all from the one plot. To do this, we can roll our own color vector...

```{r}
#placeholder color vector
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] <- "pink"
mycols[row.names(candy)=="Milky Way Midnight"] <- "blue"

# mycols
```


```{r}
ggplot(candy) +
  aes(x = winpercent, 
      y = reorder(rownames(candy),winpercent)) +
  geom_col(fill = mycols)
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

Plot of winpercent vs pricepercent to see what would be the best candy to buy...

```{r}
mycols[as.logical(candy$fruity)] <- "red"
```


```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 8)
```


> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

```{r}
library(corrplot)

cij <- cor(candy)
cij
corrplot(cij, diag = F)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar, nougat and bar, chocolate and winpercent.

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1], pca$x[,2], col=mycols, pch=16)
```

How do the original variables (columns) contribute to the new PCs. I will lok at PC1 here.

```{r}
loadings <- as.data.frame(pca$rotation)

ggplot(loadings) +
  aes(PC1, reorder(rownames(loadings), PC1), fill=PC1) +
  geom_col()
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```


```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)

p
```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

These are fruity candies. The three categories show that PC1 positive direction corresponds to hard, fruity candies that come in multiples. This makes sense since those are most fruity candies.

